---
title: "Exercise 2 guidelines"
author: "Miquel De Cáceres"
date: '2022-06-14'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Exercise objectives

  1. Perform a basic water balance run on a real-case data and inspect the results
  2. Evaluate the performance of the water balance model with observed data
  3. Perform an advanced water balance run on the same data and inspect the results
  4. Compare the results and performance between the two models
  

## Exercise data
### Font-Blanche research forest

 + The Font-Blanche research forest, located in southeastern France (43º14'27'' N 5°40'45'' E) at 420 m elevation)
 + It is composed of a top strata of *Pinus halepensis* (Aleppo pine) reaching about 12 m, a lower strata of *Quercus ilex* (holm oak), reaching about 6 m, and an understorey strata dominated by *Quercus coccifera* and *Phillyrea latifolia*. 
 + It grows on rocky and shallow soils that have a low retention capacity and are of Jurassic limestone origin. 
 + The climate is Mediterranean, with a water stress period in summer, cold or mild winters and most precipitation occurring between September and May. 


### Target stand

+ The experimental site, which is dedicated to study forest carbon and water cycles, has an enclosed area of 80×80 m but our target stand is a quadrat of dimensions 25x25 m. 
+ The following observations are available for year 2014:

   + Stand total evapotranspiration estimated using an Eddy-covariance flux tower.
   + Soil moisture content of the topmost (0-30 cm) layer.
   + Cohort transpiration estimates derived from sapflow measurements for *Q. ilex* and *P. halepensis*.
   + Pre-dawn and midday leaf water potentials for *Q. ilex* and *P. halepensis*.

## Guidelines by steps

1. Load the file `"fb_data.rds"`, which contains all the necessary data, and inspect its elements until you understand them

```{r}

```

2. Build a `forest` object including the three tree cohorts given in the input data.

```{r}

```

3. Build an object `soil` using the physical soil description included in the input data, and examine its water holding and extractable water capacity.

```{r}

```

4. Inspect the data frame `customParams` and modify the default species parameter table provided in medfate (hint: use function `modifySpParams()`) 

```{r}

```

5. Assemble a water balance input object and examine the dates of the weather dataset included in the input data

```{r}

```

6. Run a basic water balance simulation and make sure you understand the console output

```{r}

```

7. Compare the precipitation events and the generation of surface runoff and/or deep drainage (hint: use S3 function `plot()` for objects `spwb` or function `shinyplot()`).

```{r}

```

8. Examine the evapotranspiration flows (intercepted water, plant transpiration, bare soil evaporation), their seasonality and importance for the overall evapotranspiration.

```{r}

```

9. Examine the dynamics of water potential in the different soil layers. Which ones are more dynamic? What could be wrong in the model parameterization if our expectation in this respect would not be met?

```{r}

```

10. Examine the observed (measured) data set. Make sure you understand the column names. Try plotting a comparison of the observed vs. simulated stand evapotranspiration (hint: see documentation for function `?evaluation_plot`). 

```{r}

```

11. The observed data does not have the same spikes as modelled data corresponding to precipitation events. Repeat the comparison but excluding the intercepted water (hint: see option `type = "SE+TR"` in the documentation of `evaluation_plot`).

```{r}

```

12. Generate the evaluation statistics for the evapotranspiration comparison (hint: use function `evaluation_stats()`)

```{r}

```

13. Compare the observed vs. modelled soil water content. Why are the two series not scaled? Try repeating the comparison with relativized values (hint: `type = REW`). Calculate evaluation statistics.

```{r}

```

14. Create a new water balance input object for the advanced model (i.e. use `transpirationMode = "Sperry"`) and launch a new water balance simulation (this will take 1-2 minutes).

```{r}

```

15. Compare the performance of the first simulation to the second, in terms of their fit to observed soil moisture and stand evapotranspiration data.

```{r}

```

16. Compare the soil layer moisture dynamics of the two simulations. Which model produces stronger coupling between soil layers? Why?

```{r}

```

17. Using the results of the advanced water balance model, draw the plant extraction from soil layers (`"PlantExtraction"`) and the water input from hydraulic redistribution (`"HydraulicRedistribution"`) to understand the upward and downward water flows.

```{r}

```

18. Compare the plant transpiration dynamics of the two simulations at the leaf area level. Which species starts closing its stomata earlier? Are the two models consistent in this respect? Which functional parameters govern plant transpiration dynamics?

```{r}

```

19. Evaluate the plant transpiration flows predicted by the advanced model for *Pinus halepensis* and *Quercus ilex* using observed sap-flow data. For which species does the model perform better in terms of absolute values? For which species is the correlation between observed and modelled values higher?

```{r}

```

20. Draw the daily drop in leaf water potential predicted by the advanced model for the three species in the stand (hint: use `"LeafPsiRange"`). Which species has a stronger stomatal control?

```{r}

```

21. Compare the predicted vs. observed leaf water potentials (pre-dawn and midday). Is the model  appropriately estimating leaf water potential dynamics with the amount of information available?

```{r}

```

22. Examine the seasonality of plant drought stress (as defined in the model) and stem percent loss of conductance. What are the differences between the two plant drought indicators? Are the results different depending on the water balance model? 

```{r}

```

23. What species parameters are governing transpiration and drought impacts in the two models? Are their values correct?

```{r}

```



