---
title: "Exercise 2 guidelines"
author: "Miquel De Cáceres"
date: '2022-05-16'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Exercise objectives

  1. Perform a basic water balance run on a real-case data and inspect the results
  2. Evaluate the performance of the water balance model with observed data
  3. Perform an advanced water balance run on the same data and inspect the results
  4. Compare the results and performance between the two models
  

## Exercise data
### Font-Blanche research forest

 + The Font-Blanche research forest, located in southeastern France (43º14'27'' N 5°40'45'' E) at 420 m elevation)
 + It is composed of a top strata of *Pinus halepensis* (Aleppo pine) reaching about 12 m, a lower strata of *Quercus ilex* (holm oak), reaching about 6 m, and an understorey strata dominated by *Quercus coccifera* and *Phillyrea latifolia*. 
 + It grows on rocky and shallow soils that have a low retention capacity and are of Jurassic limestone origin. 
 + The climate is Mediterranean, with a water stress period in summer, cold or mild winters and most precipitation occurring between September and May. 


### Target stand

+ The experimental site, which is dedicated to study forest carbon and water cycles, has an enclosed area of 80×80 m but our target stand is a quadrat of dimensions 25x25 m. 
+ The following observations are available for year 2014:

   + Stand total evapotranspiration estimated using an Eddy-covariance flux tower.
   + Soil moisture content of the topmost (0-30 cm) layer.
   + Cohort transpiration estimates derived from sapflow measurements for *Q. ilex* and *P. halepensis*.
   + Pre-dawn and midday leaf water potentials for *Q. ilex* and *P. halepensis*.

## Guidelines

1. Load the file `"fb_data.rds"`, which contains all the necessary data, and inspect its elements until you understand them

```{r}

```

2. Build a `forest` object including the three tree cohorts given in the input data.

```{r}

```

3. Build an object `soil` using the physical soil description included in the input data, and examine its water holding and extractable water capacity.

```{r}

```

4. Inspect the data frame `customParams` and modify the default species parameter table provided in medfate (hint: use function `modifySpParams()`) 

```{r}

```

5. Assemble a water balance input object and examine the dates of the weather dataset included in the input data
```{r}

```

6. Run a basic water balance simulation and make sure you understand the console output
```{r}

```

7. 