---
title: "1.3 - Model inputs"
author: "Miquel De Cáceres"
institute: "Ecosystem Modelling Facility, CREAF"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["resources/css/custom.css", "default-fonts", "default"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warning=FALSE}
options(
  htmltools.dir.version = FALSE,
  width = 100
)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(medfate)
library(sp)
```

class: title-slide

# I.3 - Model inputs

## Miquel De Cáceres

### Ecosystem Modelling Facility

### `r as.Date("2022-06-13")`

<img class="logo-title" src="resources/img/emf_logo.svg" width=300>

---
class: center, middle

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

<div class=my-footer></div>


# Outline

##  1. Species parameters
##  2. Forest input
##  3. Vertical profiles
##  4. Soil input
##  5. Weather input
##  6. Simulation control
##  7. Simulation input object
  
```{r, include=FALSE}
library(medfate)
library(medfateutils)
```

---

layout:true

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

<div class=my-footer></div>

# 1. Species parameters

---

Simulation models in **medfate** require a `data.frame` with species parameter values.

--

The package includes a default data set of parameter values for a set of Mediterranean species.

.code80[
```{r, echo = TRUE}
data("SpParamsMED")
```
]

--

A large number of parameters (columns) can be found in `SpParamsMED`. Not all parameters are needed for all models. You can find parameter definitions in table `SpParamsDefinition`:

.code80[
```{r, echo = TRUE}
data("SpParamsDefinition")
```
]

---

.font80[
```{r, eval= require("DT"), tidy = FALSE, echo = FALSE}
DT::datatable(SpParamsDefinition,
              fillContainer = FALSE, options=list(pageLength = 6))
```
]

---

layout:true

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

<div class=my-footer></div>


# 2. Forest input

---

## Forest object

Each forest plot is represented in an object of class `forest`, a list that contains several elements. 

.code80[
```{r, include = TRUE, echo = TRUE}
forest <- exampleforestMED
```
]

--

The most important items are two data frames, `treeData` (for trees):

.code80[
```{r, echo = TRUE}
forest$treeData
```
]

--

and `shrubData` (for shrubs):

.code80[
```{r, echo = TRUE}
forest$shrubData
```
]

---

.pull-left[
## Tree data

.font90[
| Variable | Definition |
| -------- | --------------------- |
|`Species` | Species numerical code (should match `SpIndex` in `SpParams`) |
| `N` | Percentage of clay (within volume of soil particles) |
| `DBH` | Tree diameter at breast height (in cm) |
| `Height` | Tree total height (in cm) |
| `Z50` | Soil depth corresponding to 50% of fine roots (mm) |
| `Z95` | Soil depth corresponding to 95% of fine roots (mm) |
]
]
--

.pull-right[

## Shrub data

.font90[
| Variable | Definition |
| -------- | --------------------- |
| `Species` | Species numerical code (should match `SpIndex` in `SpParams`) |
| `Cover` | Shrub cover (%)  |
| `Height` | Shrub total height (in cm) |
| `Z50` | Soil depth corresponding to 50% of fine roots (mm) |
| `Z95` | Soil depth corresponding to 95% of fine roots (mm) |
]
]

---

## Creating a 'forest' object from forest inventory data

Forest inventories can be conducted in many different ways, which means that the starting form of forest data is very diverse.

--

Building `forest` objects from inventory data will always require some manual data wrangling, but package **medfateutils** provides functions that may be helpful:

--
.font90[
|  Function | Description                 |
| --------- | --------------------------- |
| `forest_mapShrubTable()` | Helps filling `shrubData` table  |
| `forest_mapTreeTable()` | Helps filling `treeData` table  |
| `forest_mapWoodyTables()` | Helps filling a `forest` object  |
| `IFN2forestlist()` | Creates a list of `forest` objects from Spanish forest inventory data   |
]


---

## Forest attributes

The package includes a number of functions to examine properties of the plants conforming the `forest` object:

-   `plant_*`: Cohort-level information (species name, id, leaf area index, height...).
-   `species_*`: Species-level attributes (e.g. basal area, leaf area index).
-   `stand_*`: Stand-level attributes (e.g. basal area).

.pull-left[
.code80[
```{r, echo = TRUE}
plant_basalArea(forest)
```
]
]

--

.pull-right[
.code80[
```{r, echo = TRUE}
plant_LAI(forest, SpParamsMED)
```
]
]

---
## Aboveground data

An important information for simulation model is the estimation of initial leaf area index and crown dimensions for each plant cohort, which is normally done using *allometries*.

--

We can illustrate this step using function `forest2aboveground()`:

.code80[
```{r, echo = TRUE}
above <- forest2aboveground(forest, SpParamsMED)
above
```
]

where species-specific allometric coefficients are taken from `SpParamsMED`.

--

Users will not normally call `forest2aboveground()`, but is important to understand what is going on behind the scenes.

---

layout: true

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

<div class=my-footer></div>

# 3. Vertical profiles

---

## Leaf distribution

Vertical leaf area distribution (at the cohort-, species- or stand-level) can be examined using:

.pull-left[
.code70[
```{r, fig = TRUE, fig.width= 6, fig.height=4, fig.align= 'center', echo=TRUE}
vprofile_leafAreaDensity(forest, SpParamsMED)
```
]
]

--

.pull-right[
.code70[
```{r, fig = TRUE, fig.width= 6, fig.height=4, fig.align= 'center', echo=TRUE}
  vprofile_leafAreaDensity(forest, SpParamsMED, 
      byCohorts = TRUE, bySpecies = TRUE)
```
]
]

---

## Radiation extinction

Radiation extinction (PAR or SWR) profile across the vertical axis can also be examined:

.pull-left[
.code80[
```{r, fig = TRUE, fig.width= 6, fig.height=4, fig.align= 'center', echo=TRUE}
vprofile_PARExtinction(forest, SpParamsMED)
```
]
]

--

.pull-right[
.code80[
```{r, fig = TRUE, fig.width= 6, fig.height=4, fig.align= 'center', echo=TRUE}
vprofile_SWRExtinction(forest, SpParamsMED)
```
]
]

---

## Belowground root distribution

Users can visually inspect the distribution of fine roots of `forest` objects by calling function `vprofile_rootDistribution()`:

.center[
.code80[
```{r, fig = TRUE, fig.width= 6, fig.height=4, echo=TRUE}
vprofile_rootDistribution(forest, SpParamsMED)
```
]
]


---
## Interactive forest inspection

Function `shinyplot()` is a more convenient way to display properties and profiles of `forest` objects:

```{r, eval = FALSE, include = TRUE, echo = TRUE}
shinyplot(forest, SpParamsMED)
```



---

layout: true

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

<div class=my-footer></div>

# 4. Soil input

---

## Soil physical description

Soil physical characteristics are specified using a **data.frame** with soil layers in rows and attributes in columns:

   + `widths` - layer widths, in mm.
   + `clay` - Percentage of clay (within volume of soil particles). 
   + `sand` - Percentage of sand (within volume of soil particles). 
   + `om` - Percentage of organic matter per dry weight (within volume of soil particles).
   + `bd` - Bulk density (g/cm3)
   + `rfc` - Rock fragment content (in whole-soil volume).

They can be initialized to default values using function `defaultSoilParams()`:

.code80[
```{r, echo = TRUE}
spar <- defaultSoilParams(2)
print(spar)
```
]

---
## Drawing soil physical attributes from *SoilGrids*

*SoilGrids* is a global database of soil properties:

.center[
.font90[
*Hengl T, Mendes de Jesus J, Heuvelink GBM, Ruiperez Gonzalez M, Kilibarda M, Blagotic A, et al. (2017) SoilGrids250m: Global gridded soil information based on machine learning. PLoS ONE 12(2): e0169748. doi:10.1371/journal.pone.0169748.*
]
]

--

Package **medfateutils** allows retrieving Soilgrids data by connecting with the SoilGrids [REST API](https://rest.isric.org)

--

To start with, we need an object of class `SpatialPoints` (from package **sp**) containing the geographic coordinates of our target forest stand:

```{r}
cc <- matrix(c(1.32, 42.20), nrow=1)
sp <- SpatialPoints(cc, proj4string = CRS(SRS_string = "EPSG:4326"))
```

--

We then call `soilgridsParams()` along with a desired vertical width of soil layers:

```{r, eval = FALSE}
soilgridsParams(sp, widths = c(300, 700, 1000))
```


---
## Soil input object

Soil input for simulations is an object of class `soil` (a list) that is created from physical description using a function with the same name:

```{r, echo = TRUE}
examplesoil <- soil(spar)
class(examplesoil)
```

---
## Water retention curves

The **water retention curve** is used to represent the relationship between soil water content ( $\theta$ ; %) and soil water potential ( $\Psi$ ; MPa).

--

The following code calls function `soil_retentionCurvePlot()` to illustrate the difference between the two water retention models in this soil:

.center[
.code80[
```{r, fig = TRUE, fig.width= 6, fig.height=4, echo=TRUE}
soil_retentionCurvePlot(examplesoil, model="both")
```
]
]

---
layout: true

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

<div class=my-footer></div>

---

# 5. Weather input

All simulations in the package require **daily weather** inputs in form of a `data.frame` with dates as `row.names`.

--

An example of daily weather data frame: 

.code80[
```{r, echo = TRUE}
data(examplemeteo)
head(examplemeteo, 2)
```
]

--

The weather variables that are required depend on the complexity of model we are using. Simulation functions have been designed to accept data frames generated using package [meteoland](https://emf-creaf.github.io/meteoland/). 


---
# 6. Simulation control

The behaviour of simulation models can be controlled using a set of global parameters.

The default parameterization is obtained using function `defaultControl()`:

```{r, echo = TRUE}
control <- defaultControl()
```

A large number of control parameters exist:

```{r, echo = TRUE, eval = FALSE, include=TRUE}
names(control)
```


Control parameters should normally be left to their **default values** until their effect on simulations is fully understood.

---

# 7. Simulation input object

Simulation functions `spwb()` and `growth()` require combining forest, soil, species-parameter and simulation control inputs into a *single input object*.

--

The combination can be done via functions  `forest2spwbInput()` and `forest2growthInput()`:

.code80[
```{r, echo = TRUE}
{{x <- forest2spwbInput(forest, examplesoil, SpParamsMED, control)}}
```
]

--

Having this additional step is handy because initialized cohort-level parameters and state variables can then be modified before calling the actual simulation functions.

--

Function `fordyn()` is different from the other two models: the user enters forest, soil, weather, species parameters and simulation control inputs *directly* into the simulation function.

