---
title: "Exercise 4b solution"
author: "Miquel De Cáceres"
date: "2024-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercise setting


##  Objectives

  1. Learn to conduct watershed simulations using **medfateland**.
  2. Learn the importance of soil water fluxes sub-model in watershed simulations.
  3. Compare watershed outflow in static vs. dynamic watershed landscapes.

## Data

Due to the long computation times of watershed simulations, this exercise uses the example
watershed data included in the **medfateland** package. The watershed is a small catchment of 66 ha (0.66 km2) 
in Catalonia.

# Exercise solution

## Step 1. Load medfateland and example watershed data

```{r}
library(medfateland)
library(ggplot2)
```

```{r}
data("example_watershed")
data("examplemeteo")
data("SpParamsMED")
example_watershed$crop_factor <- NA
example_watershed$crop_factor[example_watershed$land_cover_type=="agriculture"] <- 0.75
b <- sf::st_bbox(example_watershed)
r <-terra::rast(xmin = 401380, ymin = 4671820, xmax = 402880, ymax = 4672620, 
                nrow = 8, ncol = 15, crs = "epsg:32631")
```


## Step 2. Initialize watershed using a multi-bucket soil model

```{r}
spwb_init_buckets <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                          model = "spwb",
                                          local_control = defaultControl(soilDomains = "buckets"),
                                          simplify = TRUE)
```

## Step 3. Run watershed water balance simulation 

```{r}
ws_control <- default_watershed_control("tetis")
ws_control$tetis_parameters$num_daily_substeps <- 4
ws_control$tetis_parameters$R_baseflow <- 1
ws_control$tetis_parameters$n_baseflow <- 0.7
ws_control$tetis_parameters$R_interflow <- 1
ws_control$tetis_parameters$n_interflow <- 0.5
ws_control$tetis_parameters$R_localflow <- 1
```


```{r}
res_buckets_1y <- spwb_land(r, spwb_init_buckets, SpParamsMED, 
                            meteo = examplemeteo, 
                            watershed_control = ws_control)
plot(res_buckets_1y$watershed_balance$WatershedExport, type="l")
```


## Step 4. Update landscape and run another year of simulation 

```{r}
spwb_burnin_buckets <- update_landscape(spwb_init_buckets, res_buckets_1y)
```

```{r}
res_buckets_2y <- spwb_land(r, spwb_burnin_buckets, SpParamsMED, 
                            examplemeteo, 
                            watershed_control = ws_control)
plot(res_buckets_2y$watershed_balance$WatershedExport, type="l")
```

## Step 5. Run yet another year and check that watershed export does not change
```{r}
spwb_burnin_buckets <- update_landscape(spwb_init_buckets, res_buckets_2y)
```

```{r}
res_buckets_3y <- spwb_land(r, spwb_burnin_buckets, SpParamsMED, examplemeteo, 
                            watershed_control = ws_control)
plot(res_buckets_3y$watershed_balance$WatershedExport, type="l")
```


## Step 6. Reinitialize watershed using single-domain soil fluxes and compare watershed outflow

```{r}
spwb_init_single <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                          model = "spwb",
                                          local_control = defaultControl(soilDomains = "single"),
                                          simplify = TRUE)
```

```{r}
res_single_1y <- spwb_land(r, spwb_init_single, SpParamsMED, examplemeteo, 
                           watershed_control = ws_control)
```

```{r}
wsb_buckets_1y <- res_buckets_1y$watershed_balance
wsb_buckets_1y[["soilDomain"]] <- "buckets"
wsb_single_1y <- res_single_1y$watershed_balance
wsb_single_1y[["soilDomain"]] <- "single"
wsb_all <- rbind(wsb_buckets_1y, wsb_single_1y)
```

```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=soilDomain))+
  theme_bw()
```

```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=Transpiration, col=soilDomain))+
  theme_bw()
```

```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=DeepDrainage, col=soilDomain))+
  theme_bw()
```

```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=CellRunoff, col=soilDomain))+
  theme_bw()
```

## Step 7. Run 6 years of watershed simulation by repeating example weather data
```{r}
meteo_4y <- rbind(examplemeteo, examplemeteo, examplemeteo, examplemeteo)
meteo_4y$dates = seq(as.Date("2001-01-01"), 
                  as.Date("2004-12-30"), by="day")
```

```{r}
res_spwb_4y <- spwb_land(r, spwb_init_buckets, SpParamsMED, 
                         meteo = meteo_4y, 
                         watershed_control = ws_control)
```

## Step 8. Run 6 year of simulations with model growth, after reinitializing for this model
```{r}
growth_init_buckets <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                            model = "growth",
                                            local_control = defaultControl(soilDomains = "buckets"),
                                            simplify = TRUE)
```

```{r}
res_growth_4y <- growth_land(r, growth_init_buckets, SpParamsMED, 
                             meteo = meteo_4y, 
                             watershed_control = ws_control)
```


## Step 9. Compare results

```{r}
wsb_spwb_4y <- res_spwb_4y$watershed_balance
wsb_spwb_4y[["model"]] <- "spwb"
wsb_growth_4y <- res_growth_4y$watershed_balance
wsb_growth_4y[["model"]] <- "growth"
wsb_all <- rbind(wsb_spwb_4y, wsb_growth_4y)
```

```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=model))+
  theme_bw()
```


```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=Transpiration, col=model))+
  theme_bw()
```



