---
title: "Exercise 4b solution"
author: "Rodrigo Balaguer"
date: '2024-11-15'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 

# Exercise setting


##  Objectives

Learn to perform forest dynamics simulations on an initialized landscape while evaluating a demand-based management scenario.

  
## Data

We will use an initialized landscape corresponding to Barcelona county (derived from exercise 4a output). Daily weather data for 2023 will be extracted through an interpolation process.

# Exercise solution

## Step 1. Load initialized landscape file

We begin by loading the packages and the initialized landscape file:

```{r, message=FALSE, warning=FALSE}
library(medfateland)
library(meteoland)
library(dplyr)
```


```{r}
bcn <- readRDS("exercises/StudentRdata/bcn.rds")
```


## Step 2. Subset and filter sf objetc

We can use the `subset()` function to create a dataframe with just wildland pixels:
```{r}
bcnFor<-subset(bcn, land_cover_type=="wildland")
```

Then we can use `filter()` function from `dplyr` package to remove lines with NULL forest objects:

```{r}
bcnFor <- bcnFor |> rowwise() |> filter(!is.null(forest))
```


## Step 3. Dominant tree species

We load `SpParamsMED` and create a new column in the `sf objetc` to specify the dominant tree specie in each forest plot applying `stand_dominantTreeSpecies()` to each row:

```{r}
data("SpParamsMED")
bcnFor$dominant_tree_species <- sapply(bcnFor$forest, stand_dominantTreeSpecies, SpParamsMED)
```

Now, we can get the dominant tree species using:

```{r}
summary(as.factor(bcnFor$dominant_tree_species))
```

## Step 4. Harvesting metrics

We first load and inspect the harvesting metrics for each target specie in the Barcelona county:

```{r}
har<-read.csv("exercises/StudentRdata/harvesting_metrics.csv")
har
```

The we load the `defaultPrescriptionsBySpecies` table and inspect it:

```{r}
data("defaultPrescriptionsBySpecies")
names(defaultPrescriptionsBySpecies)
```

Finally, we specify the management unit for each stand by first defining a column `management_unit` with missing values and then assign the corresponding row number of defaultPrescriptionsBySpecies for stands dominated by each of the target species:

```{r}
defaultPrescriptionsBySpecies$Name
```

```{r}
bcnFor$management_unit <- NA
bcnFor$management_unit[bcnFor$dominant_tree_species=="Pinus halepensis"] <- 9
bcnFor$management_unit[bcnFor$dominant_tree_species=="Pinus pinea"] <- 12
bcnFor$management_unit[bcnFor$dominant_tree_species=="Quercus ilex"] <- 19
```


## Step 5. Management scenario

We can define a management scenario object using:

```{r, eval = TRUE}
scen <- create_management_scenario(defaultPrescriptionsBySpecies, 
                                   c("Pinus halepensis"=378,"Pinus pinea"=142.5, "Quercus ilex"=52.5))
```

Then, we can inspect it:

```{r, eval = TRUE}
names(scen)
```
And find the name of the scenario type:

```{r, eval = TRUE}
scen$scenario_type
```

Finally, we add a column to set the `represented_area_ha` in the `sf object`:

```{r, eval = TRUE}
bcnFor$represented_area_ha <- 25
```



## Step 6. Meteorlogical data

We can load meteorological interpolator using:

```{r, eval = TRUE}
inter<-read_interpolator("exercises/StudentRdata/interpolator_2023_calibrated.nc")
inter
```


## Step 7. Lunch simulation

We can lunch the simulation of the management scenario by using:

```{r}
fs <- fordyn_scenario(bcnFor, SpParamsMED, meteo = inter, 
                        management_scenario = scen,
                        dispersal_control = NULL,
                        parallelize = T,
                        progress = T)
```

## Step 8. Result analyses

We can inspect result volumes during the simulation by using:

```{r}
result_1 <- fs$result_volumes
result_1
```
Also, we can inspect the volumes results by species:

```{r}
result_2 <- fs$result_volumes_spp
result_2
```

And by demand:

```{r}
result_3 <- fs$result_volumes_demand
result_3
```

Finally we can analyse different outputs at stand level by using:

```{r}
result_4 <- fs$result_sf
names(result_4)
```

```{r}
result_4$dead_tree_table[[1]]
```

**Interpretation** Approximately 185 `P.halepensis` trees died during the simulation in stand id = 4.