---
title: "Exercise 4b solution"
author: "Miquel De Cáceres"
date: "2024-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercise setting


##  Objectives

  1. Learn to conduct watershed simulations using **medfateland**.
  2. Learn the importance of soil water fluxes sub-model in watershed simulations.
  3. Compare watershed outflow in static vs. dynamic watershed landscapes.

## Data

Due to the long computation times of watershed simulations, this exercise employs the example
watershed data included in the **medfateland** package. This watershed is a small catchment of 66 ha (0.66 km2) 
in Catalonia. Most (72%) of its surface  corresponds to forests, whereas 25% corresponds to agricultural areas. Example weather dataset of **medfate** is taken as climate forcing for simplicity.

# Exercise solution

## Step 1. Load medfateland and prepare example watershed data

We begin by loading the package and the watershed sf object `example_watershed` provided:
```{r}
library(medfateland)
data("example_watershed")
example_watershed
```
We can check the land cover of different grid cells using:
```{r}
table(example_watershed$land_cover_type)
```

For agricultural areas, we need to specify a crop evapotranspiration factor, which will be used to estimate actual evapotranspiration from potential evapotranspiration. Here we simply copy the example in `?spwb_land`:
```{r}
example_watershed$crop_factor <- NA
example_watershed$crop_factor[example_watershed$land_cover_type=="agriculture"] <- 0.75
```

Watershed simulations require a raster topology. In this case, we can use the boundary box of the watershed to specify the raster limits:
```{r}
b <- sf::st_bbox(example_watershed)
b
```
And we can check coordinate spacing using:
```{r}
head(sf::st_coordinates(example_watershed))
```

Knowing the limits and that spacing between cells is 100 m, we specify the raster with function `rast()`:
```{r}
r <-terra::rast(xmin = 401380, ymin = 4671820, xmax = 402880, ymax = 4672620, 
                nrow = 8, ncol = 15, crs = "epsg:32631")
```

This is the same as shown in the example section of `?spwb_land`.

## Step 2. Initialize watershed using a multi-bucket soil model

Function `initialize_landscape()` allows initializing inputs for simulations in **medfateland**, analogously to functions `spwbInput()` or `growthInput()` from **medfate**. We learned from a previous exercise that there are three sub-models for soil water fluxes, and we are asked to use the simpler (and faster) sub-model `"buckets"`. Function `initialize_landscape()` includes the parameter `simplify` to force reducing forests to the dominant tree species and dominant shrub species, while keeping overall leaf area values. The call to `initialize_landscape()` should therefore be:

```{r}
spwb_init_buckets <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                          model = "spwb",
                                          local_control = defaultControl(soilDomains = "buckets"),
                                          simplify = TRUE)
spwb_init_buckets
```

which adds a column `state` to our data, containing objects of class `spwbInput` or `aspwbInput` (the latter in case of agricultural cells).

## Step 3. Define watershed control parameters

Watershed control parameters are initialized using function `default_watershed_control()` (analogous to `defaultControl()` in **medfate**). TETIS is the only option for lateral water fluxes at the moment. 

```{r}
ws_control <- default_watershed_control("tetis")
```

Once initialized, we can manually change the parameters regulating subsurface flows (multipliers of base conductivity values and exponents regulating responsiveness of conductivity to moisture content).

```{r}
ws_control$tetis_parameters$R_baseflow <- 1
ws_control$tetis_parameters$n_baseflow <- 0.7
ws_control$tetis_parameters$R_interflow <- 1
ws_control$tetis_parameters$n_interflow <- 0.5
ws_control$tetis_parameters$R_localflow <- 1
```

## Step 4. Run watershed water balance simulation 

Here we take the example weather data set in **medfate** to run a 1-yr simulation with function `spwb_land()`. We also need to supply a species parameter table in case some (or all cells) have not been initialized previously:

```{r}
res_buckets_1y <- spwb_land(r, spwb_init_buckets, SpParamsMED, 
                            meteo = examplemeteo, 
                            watershed_control = ws_control)
```

We can display the watershed outflow (in mm) using **ggplot2**:

```{r, fig=TRUE, fig.width=10, fig.height=4}
library(ggplot2)
ggplot(res_buckets_1y$watershed_balance)+
  geom_line(aes(x = dates, y = WatershedExport))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```

**Interpretation**: The first months of the simulation generate very little outflow (only runoff). Aquifer exfiltration starts later on. Generally speaking, the larger the watershed, the longer the warm-up period necessary. It will also depend on the specific value of parameters regulating sub-surface fluxes.

## Step 5. Update landscape and run another year of simulation 

Function `update_landscape()` allows updating an input object with the final state of a previous simulation:

```{r}
spwb_burnin_buckets <- update_landscape(spwb_init_buckets, res_buckets_1y)
```

This updates the `state` column, as well as others. In particular, we can compare the effect of the update by inspecting aquifer water content:
```{r}
summary(spwb_init_buckets$aquifer)
summary(spwb_burnin_buckets$aquifer)
```


```{r}
res_buckets_2y <- spwb_land(r, spwb_burnin_buckets, SpParamsMED, 
                            examplemeteo, 
                            watershed_control = ws_control)
```

```{r}
wsb_1y <- res_buckets_1y$watershed_balance
wsb_1y[["simulation"]] <- "First year"
wsb_2y <- res_buckets_2y$watershed_balance
wsb_2y[["simulation"]] <- "Second year"
wsb_all <- rbind(wsb_1y, wsb_2y)
```

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=simulation))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```

## Step 6. Reinitialize watershed using single-domain soil fluxes and compare watershed outflow

```{r}
spwb_init_single <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                          model = "spwb",
                                          local_control = defaultControl(soilDomains = "single"),
                                          simplify = TRUE)
```

```{r}
res_single_1y <- spwb_land(r, spwb_init_single, SpParamsMED, examplemeteo, 
                           watershed_control = ws_control)
```

```{r}
wsb_buckets_1y <- res_buckets_1y$watershed_balance
wsb_buckets_1y[["soilDomain"]] <- "buckets"
wsb_single_1y <- res_single_1y$watershed_balance
wsb_single_1y[["soilDomain"]] <- "single"
wsb_all <- rbind(wsb_buckets_1y, wsb_single_1y)
```

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=soilDomain))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=Transpiration, col=soilDomain))+
  scale_y_continuous("Plant (woody) transpiration cell-average (mm/day)")+
  theme_bw()
```

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=DeepDrainage, col=soilDomain))+
  scale_y_continuous("Deep drainage cell-average (mm/day)")+
  theme_bw()
```

```{r, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=CellRunoff, col=soilDomain))+
  scale_y_continuous("Runoff cell-average (mm/day)")+
  theme_bw()
```

## Step 7. Run 6 years of watershed simulation by repeating example weather data
```{r}
meteo_4y <- rbind(examplemeteo, examplemeteo, examplemeteo, examplemeteo)
meteo_4y$dates = seq(as.Date("2001-01-01"), 
                  as.Date("2004-12-30"), by="day")
```

```{r}
res_spwb_4y <- spwb_land(r, spwb_init_buckets, SpParamsMED, 
                         meteo = meteo_4y, 
                         watershed_control = ws_control)
```

## Step 8. Run 6 year of simulations with model growth, after reinitializing for this model
```{r}
growth_init_buckets <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                            model = "growth",
                                            local_control = defaultControl(soilDomains = "buckets"),
                                            simplify = TRUE)
```

```{r}
res_growth_4y <- growth_land(r, growth_init_buckets, SpParamsMED, 
                             meteo = meteo_4y, 
                             watershed_control = ws_control)
```


## Step 9. Compare results

```{r}
wsb_spwb_4y <- res_spwb_4y$watershed_balance
wsb_spwb_4y[["model"]] <- "spwb"
wsb_growth_4y <- res_growth_4y$watershed_balance
wsb_growth_4y[["model"]] <- "growth"
wsb_all <- rbind(wsb_spwb_4y, wsb_growth_4y)
```

```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=model))+
  theme_bw()
```


```{r}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=Transpiration, col=model))+
  theme_bw()
```



