---
title: "Exercise 4b solution"
author: "Miquel De Cáceres"
date: "2024-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercise setting


##  Objectives

  1. Learn to conduct watershed simulations using **medfateland**.
  2. Learn the importance of soil water fluxes sub-model in watershed simulations.
  3. Compare watershed outflow in static vs. dynamic watershed landscapes.

## Data

Due to the long computation times of watershed simulations, this exercise employs the example
watershed data included in the **medfateland** package. This watershed is a small catchment of 66 ha (0.66 km2) 
in Catalonia. Most (72%) of its surface  corresponds to forests, whereas 25% corresponds to agricultural areas. Example weather dataset of **medfate** is taken as climate forcing for simplicity.

# Exercise solution

## Step 1. Load medfateland and prepare example watershed data

We begin by loading the package and the watershed sf object `example_watershed` provided:
```{r}
library(medfateland)
data("example_watershed")
example_watershed
```
We can check the land cover of different grid cells using:
```{r}
table(example_watershed$land_cover_type)
```

For agricultural areas, we need to specify a crop evapotranspiration factor, which will be used to estimate actual evapotranspiration from potential evapotranspiration. Here we simply copy the example in `?spwb_land`:
```{r}
example_watershed$crop_factor <- NA
example_watershed$crop_factor[example_watershed$land_cover_type=="agriculture"] <- 0.75
```

Watershed simulations require a raster topology. In this case, we can use the boundary box of the watershed to specify the raster limits:
```{r}
b <- sf::st_bbox(example_watershed)
b
```
And we can check coordinate spacing using:
```{r}
head(sf::st_coordinates(example_watershed))
```

Knowing the limits and that spacing between cells is 100 m, we specify the raster with function `rast()`:
```{r}
r <-terra::rast(xmin = 401380, ymin = 4671820, xmax = 402880, ymax = 4672620, 
                nrow = 8, ncol = 15, crs = "epsg:32631")
```

This is the same as shown in the example section of `?spwb_land`.

## Step 2. Initialize watershed using a multi-bucket soil model

Function `initialize_landscape()` allows initializing inputs for simulations in **medfateland**, analogously to functions `spwbInput()` or `growthInput()` from **medfate**. We learned from a previous exercise that there are three sub-models for soil water fluxes, and we are asked to use the simpler (and faster) sub-model `"buckets"`. Function `initialize_landscape()` includes the parameter `simplify` to force reducing forests to the dominant tree species and dominant shrub species, while keeping overall leaf area values. The call to `initialize_landscape()` should therefore be:

```{r}
spwb_init_buckets <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                          model = "spwb",
                                          local_control = defaultControl(soilDomains = "buckets"),
                                          simplify = TRUE)
spwb_init_buckets
```

which adds a column `state` to our data, containing objects of class `spwbInput` or `aspwbInput` (the latter in case of agricultural cells).

## Step 3. Define watershed control parameters

Watershed control parameters are initialized using function `default_watershed_control()` (analogous to `defaultControl()` in **medfate**). TETIS is the only option for lateral water fluxes at the moment. 

```{r}
ws_control <- default_watershed_control("tetis")
```

Once initialized, we can manually change the parameters regulating subsurface flows (multipliers of base conductivity values and exponents regulating responsiveness of conductivity to moisture content).

```{r}
ws_control$tetis_parameters$R_baseflow <- 1
ws_control$tetis_parameters$n_baseflow <- 0.7
ws_control$tetis_parameters$R_interflow <- 1
ws_control$tetis_parameters$n_interflow <- 0.5
ws_control$tetis_parameters$R_localflow <- 1
```

## Step 4. Run watershed water balance simulation 

Here we take the example weather data set in **medfate** to run a 1-yr simulation with function `spwb_land()`. We also need to supply a species parameter table in case some (or all cells) have not been initialized previously:

```{r}
res_buckets_1y <- spwb_land(r, spwb_init_buckets, SpParamsMED, 
                            meteo = examplemeteo, 
                            watershed_control = ws_control)
```

We can display the watershed outflow (in mm) using **ggplot2**:

```{r, fig=TRUE, fig.width=10, fig.height=4}
library(ggplot2)
ggplot(res_buckets_1y$watershed_balance)+
  geom_line(aes(x = dates, y = WatershedExport))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```

**Interpretation**: The first months of the simulation generate very little outflow (only runoff). Aquifer exfiltration starts later on. Generally speaking, the larger the watershed, the longer the warm-up period necessary. It will also depend on the specific value of parameters regulating sub-surface fluxes.

## Step 5. Update landscape and run another year of simulation 

Function `update_landscape()` allows updating an input object with the final state of a previous simulation:

```{r}
spwb_burnin_buckets <- update_landscape(spwb_init_buckets, res_buckets_1y)
```

This updates the `state` column, as well as others. In particular, we can compare the effect of the update by inspecting aquifer water content:
```{r}
summary(spwb_init_buckets$aquifer)
summary(spwb_burnin_buckets$aquifer)
```

We now perform a second year of simulation using again a call to `spwb_land()` but with the updated state of the first simulation:

```{r}
res_buckets_2y <- spwb_land(r, spwb_burnin_buckets, SpParamsMED, 
                            examplemeteo, 
                            watershed_control = ws_control)
```
We now compare the watershed outflow of the two simulations by first assembling the results in a single data frame:

```{r}
wsb_1y <- res_buckets_1y$watershed_balance
wsb_1y[["simulation"]] <- "First year"
wsb_2y <- res_buckets_2y$watershed_balance
wsb_2y[["simulation"]] <- "Second year"
wsb_all <- rbind(wsb_1y, wsb_2y)
```

and then displaying the outflows together:

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=simulation))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```
**Interpretation**: As expected, in the first part of the year the results are very different. However, by the end of the year the two simulations produce the same outflow. This is an indication that we have achieved a constancy in behavior. As indicated, larger watersheds will require longer warm-up periods.


## Step 6. Reinitialize watershed using single-domain soil fluxes and compare watershed outflow

In order to show the importance of local sub-model of bulk soil water fluxes, we re-initialize the watershed inputs, but this time specifying `soilDomains = "single"` in our local control parameters:
```{r}
spwb_init_single <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                          model = "spwb",
                                          local_control = defaultControl(soilDomains = "single"),
                                          simplify = TRUE)
```

An perform a simulation analogous to step #4:
```{r}
res_single_1y <- spwb_land(r, spwb_init_single, SpParamsMED, examplemeteo, 
                           watershed_control = ws_control)
```

To compare the results of the two simulations, we first assemble a single data frame:

```{r}
wsb_buckets_1y <- res_buckets_1y$watershed_balance
wsb_buckets_1y[["soilDomain"]] <- "buckets"
wsb_single_1y <- res_single_1y$watershed_balance
wsb_single_1y[["soilDomain"]] <- "single"
wsb_all <- rbind(wsb_buckets_1y, wsb_single_1y)
```

and plot the watershed outflow together:

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=soilDomain))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```

we can also compare the dynamics of other watershed-level average flows, such as plant transpiration, run-off and deep drainage:

```{r, fig=TRUE, fig.width=10, fig.height=11}
g1 <- ggplot(wsb_all)+
  geom_line(aes(x=dates, y=Transpiration, col=soilDomain))+
  scale_y_continuous("Plant (woody) transpiration cell-average (mm/day)")+
  ggtitle("Transpiration")+
  theme_bw()
g2 <- ggplot(wsb_all)+
  geom_line(aes(x=dates, y=DeepDrainage, col=soilDomain))+
  scale_y_continuous("Deep drainage cell-average (mm/day)")+
  ggtitle("Deep drainage")+
  theme_bw()
g3 <- ggplot(wsb_all)+
  geom_line(aes(x=dates, y=CellRunoff, col=soilDomain))+
  scale_y_continuous("Runoff cell-average (mm/day)")+
  ggtitle("Runoff")+
  theme_bw()
cowplot::plot_grid(g1, g2, g3, nrow =3)
```

**Interpretation**: The outflow hydrological response at the watershed level is strikingly different, highlighting the importance of the choice of the local sub-model for soil water fluxes. However, the two simulations do not differ much in terms of plant transpiration, which suggests that the choice is not critical when the focus is on vegetation stress and dynamics. As expected, watershed-average of cell runoff and deep drainage also differs between local sub-models.

## Step 7. Run 4 years of watershed simulation by repeating example weather data

In this last part of the exercise, the idea is to compare watershed simulations assuming a static vegetation vs. simulations where growth and mortality processes can occur. To do so, we start by simulating 4 years of watershed simulations with the `spwb_land()` function, which involves assuming a static vegetation. As weather data, we simply repeat four times the 1-year weather data, while setting the appropriate dates:

```{r}
meteo_4y <- rbind(examplemeteo, examplemeteo, examplemeteo, examplemeteo)
meteo_4y$dates = seq(as.Date("2001-01-01"), 
                  as.Date("2004-12-30"), by="day")
```

Then we call the simulation (this will take a few minutes):

```{r}
res_spwb_4y <- spwb_land(r, spwb_init_buckets, SpParamsMED, 
                         meteo = meteo_4y, 
                         watershed_control = ws_control)
```

## Step 8. Run 4 years of simulations with model growth, after initializing inputs for this model

Next, we want to run similar simulation but allowing growth and mortality to occur. We first, initialize the watershed for model `"growth"` in a call to `initialize_landscape()` where we also specify `"buckets"` as the soil water flux sub-model:

```{r}
growth_init_buckets <- initialize_landscape(example_watershed, SpParams = SpParamsMED,
                                            model = "growth",
                                            local_control = defaultControl(soilDomains = "buckets"),
                                            simplify = TRUE)
```

The call to `growth_land()` is analogous to `spwb_land()`:

```{r}
res_growth_4y <- growth_land(r, growth_init_buckets, SpParamsMED, 
                             meteo = meteo_4y, 
                             watershed_control = ws_control)
```

Note that the fact that we simplified vegetation into dominant cohorts make our growth dynamics less realistic, but we did so to speed up calculations.



## Step 9. Compare the results of the previous two steps

To compare the watershed outflow between `spwb` and `growth` models we assemble a data frame with all the results:

```{r}
wsb_spwb_4y <- res_spwb_4y$watershed_balance
wsb_spwb_4y[["model"]] <- "spwb"
wsb_growth_4y <- res_growth_4y$watershed_balance
wsb_growth_4y[["model"]] <- "growth"
wsb_all <- rbind(wsb_spwb_4y, wsb_growth_4y)
```

and then draw the outflow dynamics using:

```{r, fig=TRUE, fig.width=10, fig.height=4}
ggplot(wsb_all)+
  geom_line(aes(x=dates, y=WatershedExport, col=model))+
  scale_y_continuous("Watershed outflow (mm/day)")+
  theme_bw()
```

It is clear that the two simulations diverge progressively. Before interpreting the result, we can plot other water balance components, like watershed-averages of transpiration, runoff and deep drainage:

```{r, fig=TRUE, fig.width=10, fig.height=11}
g1 <- ggplot(wsb_all)+
  geom_line(aes(x=dates, y=Transpiration, col=model))+
  scale_y_continuous("Plant (woody) transpiration cell-average (mm/day)")+
  ggtitle("Transpiration")+
  theme_bw()
g2 <- ggplot(wsb_all)+
  geom_line(aes(x=dates, y=DeepDrainage, col=model))+
  scale_y_continuous("Deep drainage cell-average (mm/day)")+
  ggtitle("Deep drainage")+
  theme_bw()
g3 <- ggplot(wsb_all)+
  geom_line(aes(x=dates, y=CellRunoff, col=model))+
  scale_y_continuous("Runoff cell-average (mm/day)")+
  ggtitle("Runoff")+
  theme_bw()
cowplot::plot_grid(g1, g2, g3, nrow =3)
```

**Interpretation**: The simulation with growth seems to have led to a decrease in leaf area index across the watershed during the first two years, as a result of either mortality or defoliation processes. This is clear after inspecting the decrease in transpiration, which leads to higher soil moisture levels and, therefore, higher deep drainage. After the initial two years, the vegetation in the watershed seems to be at equilibrium with the climatic input and differences are simply maintained.

