---
title: "Exercise 1 solution"
author: "Miquel De Cáceres"
date: '2022-06-13'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 


# Exercise setting

## Objectives

1. Build `forest` objects from a tree data frame of forest inventory data
2. Retrieve soil physical properties from SoilGrids
3. Interpolate daily weather on the plot location

## Data

Package **medfate** includes a data frame (`poblet_trees`), corresponding to forest inventory data in a dense holm oak forest. 

   + *Location*: Poblet (Catalonia, Spain); long/lat: 1.0219º, 41.3443º
   + *Topography*: elevation = 850 m, slope = 15.1º, aspect = 15º
   + *Plot*: Circular plot of 15-m radius
   + *Tree data*: Stem diameter measurements on two plots: *control* and *managed*. 

As a result of the abandonment of former coppicing in the area, there is a high density of stems per individual in the control plot. 

The management involved a reduction of the number of stems per individual (*sucker cutback* or *selecció de tanys*). 


# Exercise solution

## Step 1. Loading packages

We begin by loading packages **medfate**, **medfateland** and **meteoland** into the search path: 

```{r, eval = TRUE, message=FALSE}
library(medfate)
library(medfateland)
library(meteoland)
```

## Step 2. Load and inspect Poblet data

Normally, tree data would be in a **.csv** or **.xlsx** file. Here, we simply load the tree data from Poblet included in the package:

```{r}
data("poblet_trees")
```

We can inspect its content, for example using:

```{r}
summary(poblet_trees)
```

The data frame includes tree data corresponding to three forest inventories:

```{r}
table(poblet_trees$Plot.Code)
```

## Step 3. Mapping trees from the control stand

We initialize an empty forest object using function `emptyforest()` from package **medfate**:

```{r, echo = TRUE}
pobl_ctl <- emptyforest()
```

To fill data for element `treeData` in the `forest` object, we need to define a mapping from column names in `poblet_trees` to variables in `treeData`. The mapping can be defined using a **named string vector**, i.e. a vector where element names are variable names in `treeData` and vector elements are strings of the variable names in `poblet_trees`:

```{r}
mapping <- c("Species.name" = "Species", "DBH" = "Diameter.cm")
```

*Note*: Using `"Species.name" = "Species"` we indicate that the function should interpret values in column `Species` as species names, not species codes.

We can now replace the empty `treeData` in `pobl_ctl` using functions `dplyr::filter()` and `forest_mapTreeTable()`:

```{r}
pobl_ctl$treeData <- poblet_trees |>
  dplyr::filter(Plot.Code=="POBL_CTL") |>
  forest_mapTreeTable(mapping_x = mapping, SpParams = SpParamsMED)
```

## Step 4. Check the mapping result

We can inspect the result using:

```{r}
summary(pobl_ctl$treeData)
```

One way to evaluate if the tree data is correctly specified is to display a summary of the `forest` object using the `summary()` function defined in **medfate** for this object class:

```{r}
summary(pobl_ctl, SpParamsMED)
```


Are the values of tree density, stand basal area and stand LAI acceptable for a dense oak forest?

## Step 5. Specifying plot size

We were told that forest stand sampling was done using a *circular plot* whose radius was 15 m. We can calculate the sampled area using:

```{r}
sampled_area <- pi*15^2
```

and use this information to map the tree data again, where we specify the parameter `plot_size_x`:

```{r}
pobl_ctl$treeData <- poblet_trees |>
  dplyr::filter(Plot.Code=="POBL_CTL") |>
  forest_mapTreeTable(mapping_x = mapping, SpParams = SpParamsMED,
                      plot_size_x = sampled_area)
```

We run again the summary and check whether stand's basal area and LAI make more sense:

```{r, eval = TRUE}
summary(pobl_ctl, SpParamsMED)
```

## Step 6. Adding tree heights

Leaf area density profiles describe the vertical variation of leaf area density across the canopy. In **medfate** they can be estimated using function `vprofile_leafAreaDensity()`:

```{r, eval = FALSE}
vprofile_leafAreaDensity(pobl_ctl, SpParamsMED)
```

If you try the above code it will produce an error. This is because tree heights are missing.

We should somehow estimate tree heights (in cm), for example using an allometric relationship:

```{r}
poblet_trees$Height.cm <- 100 * 1.806*poblet_trees$Diameter.cm^0.518
summary(poblet_trees$Height.cm)
```

Once tree heights are defined, we can include them in our mapping vector:

```{r}
mapping <- c("Species.name" = "Species", "DBH" = "Diameter.cm", "Height" = "Height.cm")
```

and rerun the tree data mapping.

```{r, include = FALSE}
pobl_ctl$treeData <- poblet_trees |>
  dplyr::filter(Plot.Code=="POBL_CTL") |>
  forest_mapTreeTable(mapping_x = mapping, SpParams = SpParamsMED,
                      plot_size_x = sampled_area)
```

We can now safely evaluate the leaf area density profile:

```{r, eval = TRUE}
vprofile_leafAreaDensity(pobl_ctl, SpParamsMED)
```

## Step 7. Mapping trees from the managed stand

Now we can address the *managed* stand, which has two codes corresponding to *before* and *after* the thinning intervention. Let us first deal with the pre-thinning state:

```{r}
pobl_thi_bef <- emptyforest()
pobl_thi_bef$treeData <- poblet_trees |>
  dplyr::filter(Plot.Code=="POBL_THI_BEF") |>
  forest_mapTreeTable(mapping_x = mapping, SpParams = SpParamsMED,
                      plot_size_x = sampled_area)
summary(pobl_thi_bef$treeData)
```

Beware of the missing values in column `Species`

## Step 8. Fixing species nomenclature

The `Species` variable contains two missing values. This will normally happen when some species cannot be identified. We can verify if this happens for other parts of the Poblet tree data:

```{r}
sum(!(poblet_trees$Species %in% SpParamsMED$Name))
```

If we display species counts we can identify which species is not being parsed:

```{r}
table(poblet_trees$Species)
```

In this case, the name used for the downy oak (*Quercus humilis*) is a synonym and needs to be replaced by its accepted name (*Quercus pubescens*), e.g.:

```{r}
poblet_trees$Species[poblet_trees$Species=="Quercus humilis"] <- "Quercus pubescens"
```

## Step 8. Fixing species nomenclature

Now we repeat our mapping and check the results:

```{r, eval = TRUE}
pobl_thi_bef$treeData <- poblet_trees |>
  dplyr::filter(Plot.Code=="POBL_THI_BEF") |>
  forest_mapTreeTable(mapping_x = mapping, SpParams = SpParamsMED,
                      plot_size_x = sampled_area)
summary(pobl_thi_bef, SpParamsMED)
```

Like the control plot, the `summary()` indicates a dense oak forest.

## Step 9. Mapping trees from the managed stand

We can finally map tree data for the forest plot *after* the thinning intervention:

```{r, eval = TRUE}
pobl_thi_aft <- emptyforest()
pobl_thi_aft$treeData <- poblet_trees |>
  dplyr::filter(Plot.Code=="POBL_THI_AFT") |>
  forest_mapTreeTable(mapping_x = mapping, SpParams = SpParamsMED,
                      plot_size_x = sampled_area)
summary(pobl_thi_aft, SpParamsMED)
```

And check the reduction caused by the thinning.


## Step 10. Checking the number of cohorts

So far we have considered that each tree record should correspond to a woody cohort. We can check the number of tree cohorts in each `forest` structure using:

```{r}
nrow(pobl_ctl$treeData)
nrow(pobl_thi_bef$treeData)
nrow(pobl_thi_aft$treeData)
```

This large amount of cohorts can slow done simulations considerably!

## Step 11. Reducing the number of cohorts

One way of reducing the number of cohorts is via function `forest_mergeTrees()` from package **medfate**:

```{r}
pobl_ctl <- forest_mergeTrees(pobl_ctl)
pobl_thi_bef <- forest_mergeTrees(pobl_thi_bef)
pobl_thi_aft <- forest_mergeTrees(pobl_thi_aft)
```

By default, the function will pool tree cohorts of the same species and diameter class (defined every 5 cm). 

We can check the new number of tree cohorts using again:

```{r}
nrow(pobl_ctl$treeData)
nrow(pobl_thi_bef$treeData)
nrow(pobl_thi_aft$treeData)
```

## Step 11. Reducing the number of cohorts

We can check whether stand properties were altered using the `summary()` function:

```{r}
summary(pobl_thi_aft, SpParamsMED)
```

Function `forest_mergeTrees()` will preserve the stand density and basal area that the stand description had before merging cohorts. Other properties like leaf area index may be slightly modified.

*Tip*: It is advisable to reduce the number of woody cohorts before running simulation models in **medfate**.

## Steps 12-13. Retrieving SoilGrids data

Retrieval of soil properties from SoilGrids can be done using function `add_soilgrids()` from package **medfateland**. 

Assuming we know the plot coordinates, we first create an object `sf` (see package **sf**):

```{r}
cc <- c(1.0219, 41.3443)
coords_sf <- sf::st_sf(geometry = sf::st_sfc(sf::st_point(cc), crs = 4326))
```

This object can be used to query SoilGrids using `add_soilgrids()`:

```{r, eval = TRUE}
pobl_soil_sf <- medfateland::add_soilgrids(coords_sf, widths = c(300, 700, 1000))
```

This function returns an `sf` object with soil properties for each coordinate:

```{r}
pobl_soil_sf
```

Since we only have one coordinate, we can access its soil using:
```{r}
pobl_soil_props <- pobl_soil_sf$soil[[1]]
pobl_soil_props
```


## Steps 14-15. Building the initialized soil object

This data frame is a physical description of the soil. Remember that the initialized soil data structure for **medfate** simulations is built using function `soil()`:

```{r}
pobl_soil <- soil(pobl_soil_props)
```

We can inspect the soil definition using `summary()`.
```{r}
summary(pobl_soil)
```


*SoilGrids* usually underestimates the amount of rocks in the soil, because soil samples do not normally contain large stones or blocks. Realistic simulations should reduce the soil water holding capacity by increasing `rfc`. For example, here we will assume that the third layer contains 80% of rocks:

```{r}
pobl_soil_props$rfc[3] = 80
```

If we rebuild the soil object and inspect its properties...

```{r, eval= FALSE}
pobl_soil <- soil(pobl_soil_props)
summary(pobl_soil)
```

...we will see the decrease in water-holding capacity.


## Steps 16-17. Interpolating weather

Obtaining daily weather data suitable for simulations is not straightforward either. Here we illustrate one way of obtaining such data with package **meteoland**. 

We begin by building an object of class `sf` containing both the coordinates of our site as well as topographic variables.

```{r}
pobl_sf <- coords_sf |>
  dplyr::mutate(elevation = 850, slope = 15.1, aspect = 15)
pobl_sf
```

The more difficult part of using package **meteoland** is to assemble weather data from surface weather stations into an object of class `stars`.


Here we will assume that such an object is already available, by using the example object provided in the **meteoland** package.

```{r}
data("meteoland_interpolator_example")
```

Once we have an interpolator, obtaining interpolated weather for a set of target points is rather straightforward using function `interpolate_data()` from **meteoland**:

```{r, warning = TRUE}
pobl_meteo <- pobl_sf |>
  meteoland::interpolate_data(meteoland_interpolator_example, verbose = FALSE)
```


The output of function `interpolate_data()` is also an `sf` object, with a new column `interpolated_data`: 

```{r, warning = TRUE}
pobl_meteo
```

We can access the weather data frame by subsetting the appropriate element of `interpolated_data`:

```{r, eval = TRUE}
pobl_weather <- pobl_meteo$interpolated_data[[1]]
head(pobl_weather, 2)
```

