---
title: "Exercise 2b solution"
author: "Miquel De Cáceres"
date: '2024-11-07'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise setting

## Overall goal

Learn how sub-model of bulk soil water movement influences water balance results.

## Exercise material

+ Exercise_2b.Rmd
+ fontblanche.rds

## Font-Blanche research forest

 + The Font-Blanche research forest is located in southeastern France (43º14'27'' N 5°40'45'' E) at 420 m elevation)
 + The stand is composed of a top strata of *Pinus halepensis* (Aleppo pine) reaching about 12 m, a lower strata of *Quercus ilex* (holm oak), reaching about 6 m, and an understorey strata dominated by *Quercus coccifera* and *Phillyrea latifolia*. 
 + Soils are shallow and rocky, have a low retention capacity and are of Jurassic limestone origin. 
 + The climate is Mediterranean, with a water stress period in summer, cold or mild winters and most precipitation occurring between September and May. 

## Target stand

+ The experimental site, which is dedicated to study forest carbon and water cycles, has an enclosed area of 80×80 m but our target stand is a quadrat of dimensions 25×25 m. 
+ The following observations are available for year 2014:
   + Soil moisture content of the topmost (0-30 cm) layer.

# Exercise solution

## Step 1. Load Font-Blanche data

```{r, include=FALSE}
library(medfate)
library(ggplot2)
library(cowplot)
```

We are given all the necessary data, bundled in a single list:

```{r}
fb <- readRDS("exercises/StudentRdata/fontblanche.rds")
```

## Step 2. Build forest object

Element `fb$treeData` is already in the appropriate format for `forest` objects. Therefore, we can easily plug the tree data into a forest object:

```{r}
fb_forest <- emptyforest() 
fb_forest$treeData <- fb$treeData
```

## Step 3. Initialize soil object

A data frame with soil physical attributes are defined in `fb$soilData` but initialize a `soil` object to complete the hydraulic parameters:

```{r}
fb_soil <- soil(fb$soilData)
```

From which we can estimate the water holding capacity and extractable (down to -5 MPa) water capacity for each layer (in mm) using functions `soil_waterFC()` and `soil_waterExtractable()`:

```{r}
soil_waterFC(fb_soil)
soil_waterExtractable(fb_soil)
```

So the sums would be:
```{r}
sum(soil_waterFC(fb_soil))
sum(soil_waterExtractable(fb_soil))
```

The same information can also be found in the output of `summary()`.

```{r}
summary(fb_soil)
```

## Step 4. Species parameters

In the case of FontBlanche there is a table of preferred values for some parameters:

```{r}
fb$customParams
```

We can use function `modifySpParams()` to replace the values of parameters for the desired traits, leaving the rest unaltered:

```{r}
fb_SpParams <- modifySpParams(SpParamsMED, fb$customParams)
```

## Step 5. Initialize water balance inputs under different sub-models of vertical soil water fluxes

Since we are about to run a basic water balance simulation, we initialize `spwbInput` while using `transpirationMode = "Granier"` in the control parameters. However, in this case we also specify the option `soilDomains`, for which we have three different alternatives:

```{r}
x_buckets <- spwbInput(fb_forest, fb_soil, fb_SpParams, 
                       control = defaultControl(transpirationMode = "Granier", soilDomains = "buckets"))
x_single <- spwbInput(fb_forest, fb_soil, fb_SpParams, 
                      control = defaultControl(transpirationMode = "Granier", soilDomains = "single"))
x_dual <- spwbInput(fb_forest, fb_soil, fb_SpParams, 
                    control = defaultControl(transpirationMode = "Granier", soilDomains = "dual"))
```

## Step 6. Run water balance model while assessing computation times

Computation time can be limiting in some cases. To estimate computation time, we can insert the simulation call to `spwb()` within a block inside a call to `system.time()`, i.e.:

```{r, eval=TRUE}
system.time({fb_buckets <- spwb(x_buckets, fb$meteo, elevation = 420, latitude = 43.24)})
system.time({fb_single <- spwb(x_single, fb$meteo, elevation = 420, latitude = 43.24)})
system.time({fb_dual <- spwb(x_dual, fb$meteo, elevation = 420, latitude = 43.24)})
```

**Interpretation**: The bucket model is the fastest, followed by the single domain. The dual domain sub-model is the slowest of all three. These differences in computation time arise from the need to solve soil fluxes at fine temporal steps.

## Step 7. Compare water balance components

We can use the `summary()` function for `spwb` objects to see the annual-level values of water balance components. Since only one year is simulated, we bind the three vectors corresponding to the different sub-models as rows:

```{r}
m <- rbind(summary(fb_buckets),
           summary(fb_single),
           summary(fb_dual))
rownames(m) <- c("buckets", "single", "dual")
m
```

**Interpretation**: Rainfall inputs and intercepted water are equal regardless of the sub-model for soil fluxes.

+ The multi-bucket sub-model is the one with highest soil infiltration (hence, lowest infiltration excess), but it also predicts highest saturation excess, so the final runoff values are highest (`r round(sum(fb_buckets$WaterBalance$Runoff))`) and deep drainage values are lowest (`r round(sum(fb_buckets$WaterBalance$DeepDrainage))`).
+ The single-domain sub-model has lower infiltration, but also generates saturation excess, resulting in a moderate runoff (`r round(sum(fb_single$WaterBalance$Runoff))`). Deep drainage is also intermediate (`r round(sum(fb_single$WaterBalance$DeepDrainage))`). In addition, the model presents numerical solution problems which are indicated by non-negligible values of capillarity rise.
+ The dual-permeability sub-model is the one with lowest runoff (`r round(sum(fb_dual$WaterBalance$Runoff))`), because, despite lowest infiltration rates, has also lowest saturation excess. Percolation through macropores in this model creates highest rates of deep drainage (`r round(sum(fb_dual$WaterBalance$DeepDrainage))`). 

## Step 8. Compare runoff and deep drainage temporal patterns

We can use function `plot()` to draw precipitation input and, then, the water components contributing to water export under each sub-model (using `type = "Export"`).

```{r echo=TRUE, fig.height=9, fig.width=11, fig=TRUE}
g0 <- plot(fb_buckets)+
  scale_x_date(date_breaks = "1 month", date_labels = "%m")+
  theme(legend.position = "none")
g1 <- plot(fb_buckets, type = "Export")+
  scale_x_date(date_breaks = "1 month", date_labels = "%m")+
  scale_y_continuous(limits = c(0,60))+
  labs(title="multi-bucket")+
  theme(legend.position = "inside", legend.position.inside = c(0.35,0.60))
g2 <- plot(fb_single, type = "Export")+
  scale_x_date(date_breaks = "1 month", date_labels = "%m")+
  scale_y_continuous(limits = c(0,60))+
  labs(title="single domain")+
  theme(legend.position = "none")
g3 <- plot(fb_dual, type = "Export")+
  scale_x_date(date_breaks = "1 month", date_labels = "%m")+
  scale_y_continuous(limits = c(0,60))+
  labs(title="dual permeability")+
  theme(legend.position = "none")
plot_grid(g0, g1, g2, g3, ncol=1, rel_heights = c(0.5,1,1,1))
```

**Interpretation**: Surface run-off occurs the same day as precipitation events and, as seen in the step before, is highest under the multi-bucket model. Under this model, low values of deep drainage can last for some days after the event. The single domain and dual permeability sub-models generate a more realistic pattern of deep drainage after the event, although values are higher under the dual permeability sub-model. This is because the single-domain sub-model underestimates water flows near saturation (Richards equations are most appropriate for unsaturated soils).

## Step 9. Compare soil water content dynamics against observed data

As known from exercise 2a, observed soil moisture dynamics is in `fb$measuredData`. We can compare observed versus simulated soil moisture dynamics using function `evaluation_plot`. Here we combine plots of temporal series with scatter plots of observed vs. predicted values:

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
g1a <- evaluation_plot(fb_buckets, fb$measuredData, type = "SWC", plotType="dynamics")+
  labs(title="multi-bucket")+
  theme(legend.position = "none")
g1b <- evaluation_plot(fb_buckets, fb$measuredData, type = "SWC", plotType="scatter")+
  labs(title="")
g1 <- plot_grid(g1a, g1b, ncol = 2, rel_widths = c(1, 0.4))

g2a <- evaluation_plot(fb_single, fb$measuredData, type = "SWC", plotType="dynamics")+
  labs(title="single domain")+
  theme(legend.position = "none")
g2b <- evaluation_plot(fb_single, fb$measuredData, type = "SWC", plotType="scatter")+
  labs(title="")
g2 <- plot_grid(g2a, g2b, ncol = 2, rel_widths = c(1, 0.4))

g3a <- evaluation_plot(fb_dual, fb$measuredData, type = "SWC", plotType="dynamics")+
  labs(title="dual permeability")+
  theme(legend.position = "none")
g3b <- evaluation_plot(fb_dual, fb$measuredData, type = "SWC", plotType="scatter")+
  labs(title="")
g3 <- plot_grid(g3a, g3b, ncol = 2, rel_widths = c(1, 0.4))

plot_grid(g1, g2, g3, ncol = 1)
```

**Interpretation**: In all cases, the observed series are not scaled with simulated ones, for reasons discussed in exercise 2a. However, here we focus on the behavior of the model near saturation, where bulk soil fluxes are more relevant. Note that the dual permeability model is the one where the shape of post-event decrease in topsoil moisture is more realistic, despite the bias with observed values.
