---
title: "Exercise 4b guidelines"
author: "Miquel De Cáceres"
date: '2024-11-13'
output:
  html_document: default
---

# Exercise setting


##  Objectives

  1. Learn to conduct watershed simulations using **medfateland**.
  2. Learn the importance of soil water fluxes sub-model in watershed simulations.
  3. Compare watershed outflow in static vs. dynamic watershed landscapes.

## Data

Due to the long computation times of watershed simulations, this exercise uses the example
watershed data included in the **medfateland** package. The watershed is a small catchment of 66 ha (0.66 km2) 
in Catalonia. Example weather dataset of **medfate** is taken as climate forcing for simplicity.

## Guidelines

1. Load package **medfateland** and the example watershed provided in the package (i.e. `watershed_example`). Indicate crop evapotranspiration factor for `agriculture` cells and create the raster definition that includes one stand per cell (**hint**: see help of `?spwb_land`).

```{r}

```

2. Initialize inputs for soil plant water balance simulations on the watershed (**hint**: use function `initialize_landscape()`). For the initialization, you should: (a) specify that local soil water fluxes will be simulated using a multi-bucket model; and (b) force a simplification of the landscape to simulate only dominant species (**hint**: see option `simplify`). 

```{r}

```

3. Initialize watershed control parameters for TETIS (the only functional model). Specify the following parameter values (you can check their meaning in the medfatebook):

| Parameter                 | Value    |
|---------------------------|----------|
| `R_baseflow`              |    1     |
| `n_baseflow`              |    0.7     |
| `R_interflow`              |    1     |
| `n_baseflow`              |    0.5     |
| `R_localflow`              |    1     |

```{r}

```

4. Run a water balance simulation over the watershed for one year, using the `examplemeteo` weather data provided in package **medfate**. Examine the predicted daily watershed outflow.

```{r}

```

