---
title: "Exercise 2c guidelines"
author: "Rodrigo Balaguer"
date: '2024-11-13'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Exercise objectives

  1. Perform an advanced water balance run on a real-case data and inspect live fuel moisture content (LFMC) results.
  2. Evaluate the performance of the water balance model with observed data.
  3. Compare LFMC dynamics with annual fire hazard values.

## Exercise data
### Tivissa study plot 

 + Tivissa study plot is located in southern Tarragona mountains (41°03'10.8" N; 0°47'57.2" E) at 555 m elevation.
 + The climate is Mediterranean, with a water stress period in summer.
 + The study plot is dedicated to measure live fuel moisture content (LFMC) dynamics, which is the moisture content of live plant tissues, a variable used to predict wildfire danger.

### Target stand
 + The stand is dominated by shrublands covered with *Quercus coccifera* (30%) and *Rosmarinus officinalis* (50%).
 + The LFMC observations are available for year 2014.

## Guidelines by steps

1. Load the file `"tivissa.rds"`, which contains all the necessary data.

```{r}

```

2. Build a `forest` object including the two shrub cohorts given in the input data (*hint*: `forest` object can contain both, tree and shrub data).

```{r}

```

3. Initialize an object `soil` using the physical soil description included in the input data.

```{r}

```

4. Load the default species parameter table provided in medfate.

```{r}

```

5. Initialize simulation control parameters to run an advanced water balance simulation applying `Sureau` model. Then modify default values to display Fire Hazard results in the simulation output (*hint*: set `fireHazardResults <- TRUE`). Finally, set a standard wind speed (m/s) value for the entire simulation in order to control the effect of this variable over Fire Hazard results (*hint*: set `fireHazardStandardWind <- 5`).

```{r}

```

6. Try to initialize water balance input. Why is returning an error?

```{r}

```

7. Correct the species names in `shrubData` and repeat the water balance input initialization (*hint*: `Rosmarinus officinalis` = `Salvia rosmarinus`).

```{r}

```

8. Extract meteorological and topographic data from the input file and run the advanced water balance simulation.

```{r}

```

9. Visualize and compare both species LFMC dynamics (*hint*: use `plot()` function with `type = "LFMC"`). What could be driving the differences among both species LFMC annual patterns?

```{r}

```

10. Compare the observed vs. modeled LFMC values and describe both species performance (*hint*: `evaluation_plot()` function does not work with `LFMC`, instead we can use `summary()` function for `spwb` objects and `ggplot()`).

```{r}

```

11. Extract daily fire hazard values from the `spwb` object. Do you recognize any fire behavior variable represented in the columns? Compare precipitation, LFMC and surface fire line intensity. Which relations did you observe? (*hint*: use `summary()` function to obtain `FireHazard` results. Surface fire line intensity is named as `I_b_surface`). 

```{r}

```

